Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-06-09 SIR_manuscript.tex
more BMB comment integration

The low-testing case \fref{pan} reflects the case where $\rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (this is approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.
  The critical contour of $\Delta=1-\frac{\gamma}{\beta}$, corresponding to the threshold $\Rnum=1$, is shown as a dotted line in \fref{pan} and \fref{pan2}; combinations of $\rho$ and $\omega$ that lie above this critical contour will control the epidemic. The implied baseline value of $\Rnum=\frac{\beta}{\gamma}$ in these figures is 3, corresponding to a case where the epidemic spreads slowly even in the absence of testing and tracing.

@@

\begin{itemize}
\item We would generally expect increasing testing delays to increase \Rnum, thus decreasing effectiveness of control $\Delta$. This is in fact what happens when waiting individuals do not isolate ($\theta_w =0$, top row of \fref{pan}) --- as we move to the right within each plot in this row, $\Delta$ decreases.
However, when waiting individuals isolate ($\theta_w>0$), we more often see the opposite effect: longer testing delays lead to a greater control effect $\Delta$. This is because people waiting for negative tests are assumed to continue to isolate; this applies both to susceptibles and to people who became infected while waiting for negative test results. This effect outweighs the effect of confirmed individuals isolating except when this isolation parameter $\theta_c$ is substantially bigger than $\theta_w$. This result depends on the idea that, all else equal, people who have to wait longer for test results isolate at the same level (but for a longer time) as they would if the wait were shorter.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, however, at very high testing intensities. This can only occur when testing is targeted, and $\theta_w$ is relatively small (\fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.

@@

Over a broad range of parameter space (for random testing, approximately $\theta_w \geq 0.25$; for targeted testing, $\theta_w \geq 0.25$ and either $\theta_c \ge 0.5$ or $1/\omega > 5$), decreasing $\omega$ --- i.e., decreasing the rate at which tested people find out about their infection status --- decreases $\Rnum$. This result is counterintuitive; public health agencies have invested a lot of effort in increasing this rate. Dynamically, this effect occurs because speeding up test returns shortens the isolation period of uninfected individuals. For infected people it only shortens the time to progression to the isolation level of the confirmed-positive compartment. Slowing test returns increases $\Rnum$ only if the proportion of infectives in the tested population (test positivity), and the magnitude of increase in isolation effectiveness from the waiting to the confirmed-positive compartment, are large enough to outweigh the decrease in isolation of uninfected waiting individuals.

@@

The other counterintuitive result from our analysis is that, for sufficiently high testing intensity $\rho$, increasing testing intensity can actually increase $\Rnum$ (e.g. \fref{pan2}(b), upper right panel [$\theta_c=1$, $\theta_w=0$]). This phenomenon occurs mainly because of a peculiar feature of the DFE in our model: we assume that testing is ongoing before the beginning of the epidemic, so that the starting point of the epidemic invasion is an equilibrium distribution of susceptibles between the $S\_n$ (untested) and $S\_u$ (waiting) compartments: in particular, $S\_u^* = \rho/(\rho + \omega)$. (In reality, testing intensity itself is a dynamic variable that will increase as the epidemic proceeds.) If isolation in the waiting compartment is low ($\theta_w \ll 1$), then waiting individuals can easily be infected once the epidemic starts. Once infected they can still infect others while in the waiting compartment, and cannot be tested again before they have returned to the untested compartment. Thus, under certain extreme conditions (low $\theta_w$, high $\theta_c$, low $\omega$, high $\rho$), $\Rnum$ becomes an increasing function of testing intensity. \bmb{This phenomenon occurs primarily (only??) for targeted testing, because \ldots CAN WE FIGURE OUT WHY? Targeting should be irrelevant to determining the DFE (because everyone is $S$ at that point), so what is the dynamical explanation? Is there a second order effect somewhere, e.g. testing is more effective overall (darker colours in Fig 3(b) upper-right panel than in Fig 3(a) upper panel), so we more quickly reach the point where the effect of increasing $S\_n^*$ outweighs the beneficial effects of testing?} However, this phenomenon is even more unrealistic than the possibility that slowing test returns increases $\Rnum$. It depends both on the assumption that regular testing is occurring before the epidemic starts, and on levels of testing that are unrealistically high (at least in large, general-population settings), i.e. at least 10\% of the population being tested per day.

@@

\bmb{Leftover comments:
\begin{itemize}
\item should emphasize (at some appropriate point in text) that $\Delta$ is independent of $\beta$ (and hence of $\Rnum$ if we are thinking primarily of holding $\gamma$ constant and varying $\beta$) for the small-$\rho$ (Taylor-expansion) case, see eq. ...
\item it would be good to have a few more testing refs: Bergstrom et al., there might be some other stuff we can recycle from the lab meeting review; Friston's stuff \ali{\citep{friston2021testing} OK?} deserves to be mentioned because it does have explicit testing flows too. We should try not to ignore/piss anyone off who has already done stuff
\end{itemize}
\ali{:}
\cite{bergstrom2020frequency} modeled repeated random testing of isolated populations.
\citep{friston2021testing} model the effects of self-isolation on testing and tracking. Their model is a dynamic causal model equipped with isolation state which is an extension of compartmental model SEIR. They found that there exists no testing strategy that can attenuate mortality rates, other than by deferring or delaying a second wave.
}

@@

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder\Rnum{\omega}$, which specifically depends on $\pder\Delta{\omega}$. We use
the linearization of $\Rnum$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The Taylor expansion of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega \gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S} \omega^2 p_I \theta\_c+(\frac{w_I}{w_S}+1) \gamma\omega\theta\_w +\gamma^2 \theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder\Delta{\omega} = \frac{\rho}{\omega^2 (\omega+\gamma)^2} \Big(
(p_I \frac{w_I}{w_S}\theta\_c -(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w \gamma(\theta_w-2) \omega+\theta_w \gamma^2 (\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
% \ali{I stoped here!}

>>>>

The low-testing case \fref{pan} reflects the case where $\rho$ is small relative to the population size. Specifically, $\rho \in [0,0.013]$, and test return rate $\omega\in [1/12,2]$. This testing intensity reflects realistic testing rates during the \covid pandemic, i.e. a maximum of 1.3\% of the population per day, giving a maximum of 10000 tests/day in a population of size $N_0=10^6$ (approximately four times the maximum testing rate in Ontario, Canada in mid-2021). The less realistic high-testing case \fref{pan2} is included to highlight the occurrence of non-monotonic changes in $\Rnum$ with respect to $\rho$.
In \fref{pan2} the maximum capacity of $\rho$ is larger relative to the population size, $\rho \in [0,1/5)$ and the test return rate $\omega\in [1/5,2]$; these values are clearly unrealistic for a large population but might be relevant for some population undergoing focused testing, such as a sports league or university.

@@

\begin{itemize}
\item We would generally expect increasing testing delays to increase \Rnum, thus decreasing effectiveness of control $\Delta$. This is in fact what happens when waiting individuals do not isolate ($\theta_w =0$, top row of \fref{pan}) --- as we move to the right within each plot in this row, $\Delta$ decreases.
  However, when waiting individuals isolate ($\theta_w>0$), we more often see the opposite effect: longer testing delays lead to a greater control effect $\Delta$. This is because people waiting for negative tests are assumed to continue to isolate; this applies both to susceptibles and to people who became infected while waiting for negative test results. This effect outweighs the effect of confirmed individuals isolating except when isolation efficacy of confirmed-positive individuals ($\theta_c$) is substantially larger than that of waiting individuals ($\theta_w$).

\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, however, at very high testing intensities. This can only occur when testing is targeted, and $\theta_w$ is relatively small (\fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.

@@

Over a broad range of parameter space (for random testing, approximately $\theta_w \geq 0.25$; for targeted testing, $\theta_w \geq 0.25$ and either $\theta_c \ge 0.5$ or $1/\omega > 5$), decreasing $\omega$ --- i.e., decreasing the rate at which tested people find out about their infection status --- decreases $\Rnum$. For low testing rates, the value of $\Delta$ is independent of $\beta$ (Eq.~\ref{eq:lin}), so the patterns shown in \fref{pan} hold for any baseline values of $\Rnum$ (assuming we hold $\gamma$ constant).

This result is counterintuitive; public health agencies have invested a lot of effort in increasing this rate. Dynamically, this effect occurs because speeding up test returns shortens the isolation period of uninfected individuals. For infected people it only shortens the time to progression to the isolation level of the confirmed-positive compartment. Slowing test returns increases $\Rnum$ only if the proportion of infectives in the tested population (test positivity), and the magnitude of increase in isolation effectiveness from the waiting to the confirmed-positive compartment, are large enough to outweigh the decrease in isolation of uninfected waiting individuals.

@@

The other counterintuitive result from our analysis is that, for sufficiently high testing intensity $\rho$, increasing testing intensity can actually increase $\Rnum$ (e.g. \fref{pan2}(b), upper right panel [$\theta_c=1$, $\theta_w=0$]). This phenomenon occurs mainly because of a peculiar feature of the DFE in our model: we assume that testing is ongoing before the beginning of the epidemic, so that the starting point of the epidemic invasion is an equilibrium distribution of susceptibles between the $S\_n$ (untested) and $S\_u$ (waiting) compartments: in particular, $S\_u^* = \rho/(\rho + \omega)$. (In reality, testing intensity itself is a dynamic variable that will increase as the epidemic proceeds.) If isolation in the waiting compartment is low ($\theta_w \ll 1$), then waiting individuals can easily be infected once the epidemic starts. Once infected they can still infect others while in the waiting compartment, and cannot be tested again before they have returned to the untested compartment. Thus, under certain extreme conditions (low $\theta_w$, high $\theta_c$, low $\omega$, high $\rho$), $\Rnum$ becomes an increasing function of testing intensity. \bmb{This phenomenon occurs primarily (only??) for targeted testing, because \ldots CAN WE FIGURE OUT WHY? Targeting should be irrelevant to determining the DFE (because everyone is $S$ at that point), so what is the dynamical explanation? Is there a second order effect somewhere, e.g. testing is more effective overall (darker colours in Fig 3(b) upper-right panel than in Fig 3(a) upper panel), so we more quickly reach the point where the effect of increasing $S\_n^*$ outweighs the beneficial effects of testing?  \fady{targeting + high testing rates could exhaust the untested infectious population such that even higher testing rates (note we are asking about $\partial \Delta/\partial \rho$) could enrich the tested population even more for $S$?} This doesn't seem impossible, but checking the verbal reasoning against math seems challenging \ldots} However, this phenomenon is even more unrealistic than the possibility that slowing test returns increases $\Rnum$. It depends both on the assumption that regular testing is occurring before the epidemic starts, and on levels of testing that are unrealistically high (at least in large, general-population settings), i.e. at least 10\% of the population being tested per day.

@@

\bmb{Leftover comments:
\begin{itemize}
\item it would be good to have a few more testing refs: Bergstrom et al., there might be some other stuff we can recycle from the lab meeting review; Friston's stuff \ali{\citep{friston2021testing} OK?} deserves to be mentioned because it does have explicit testing flows too. We should try not to ignore/piss anyone off who has already done stuff
\end{itemize}
\ali{:}
Models of controlling COVID-19 spread in isolated populations (especially focused on college campuses) by
\cite{bergstrom2020frequency} (analytical) and \cite{rogers_high-frequency_2021} (computational) \ldots
(maybe other useful citations \href{https://docs.google.com/spreadsheets/d/1j1wgHNEAq_OKfOnugEen8vafXszyjw1Q}{here}?)
\citep{friston2021testing} use a compartmental model to explore the effects of self-isolation on testing and tracking in the population of the UK (they conclude that testing strategy can delay a second wave but cannot reduce overall mortality)
Blog post \citep{otto_tipping_2020} probably not citeable \ldots
}

@@

% %%%%%%%
\subsection{rate of returning tests} \label{app:omega}
{\bf The third goal} is to explore how changes in the rate of test return $\omega$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder\Rnum{\omega}$, which specifically depends on $\pder\Delta{\omega}$. We use
the linearization of $\Rnum$ around $\rho=0$ to show that there a non-monotonic relationship between $\Rnum$  and $\omega$. The first-order Taylor approximation of $\Delta$ at $\rho=0$ is
\begin{equation}
\label{eq:lin}
\Delta = \frac{\rho}{\omega \gamma(\omega+\gamma)} \Big(
\frac{w_I}{w_S} \omega^2 p_I \theta\_c+(\frac{w_I}{w_S}+1) \gamma\omega\theta\_w +\gamma^2 \theta\_w(2-\theta\_w)
\Big). 
\end{equation}
This results in
\begin{equation}
\label{eq:dlindo}
\pder\Delta{\omega} = \frac{\rho}{\omega^2 (\omega+\gamma)^2} \Big(
(p_I \frac{w_I}{w_S}\theta\_c -(1+\frac{w_I}{w_S})\theta\_w)\omega^2+2\theta_w \gamma(\theta_w-2) \omega+\theta_w \gamma^2 (\theta_w-2)
\Big),
\end{equation}
around $\rho=0$.
% \ali{I stoped here!}
